### 2022-05-25 doc/VerticalArrangements.tex
finished PRELIMINARY draft of merger sims. NEED TO CHECK THESE. And replace horrible table of sim results.
candidate replacement figure for table 3 (TrashSims). Have not yet removed table 3.
incorporates Gloria's edits for Bates White Conference.

@@

In this paper, we use a standard merger simulation model to assess the welfare implications of these complex mergers.  Our model is drawn directly from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.  We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.  In each of these simulations, we calculate the impacts on consumer welfare and firm profits.  Furthermore, we examine how the presence of additional integrated rivals in the market changes our results. Our baseline simulations assume that units are produced with constant marginal costs, but we later extend our analysis to linear upward-sloping marginal costs.

@@

@@

Turning to input prices, when wholesaler $w$ bargains with unaffiliated retailer $s$, the first order condition becomes
\begin{equation}
\begin{split}
&[p^W_{sw} - c^W_{sw}]s_{sw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^s \setminus \{w\})\\
& - \overbrace{\overbrace{[p_{rw}-c^W_{rw}-c^R_{rw}] \Delta s_{rw}(\mathbb{W}^s \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^s \setminus \{w\})}^{\text{RRC effect}}=\\
&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right).
\end{split}
\label{eq: vmerger wh upstream foc}
\end{equation}
which reflects the change in the disagreement payoff coming from the merger with retailer $r$.  Now when the wholesaler considers the possible loss of sales upon ceasing to trade with retailer $s$, these losses are softened due to a potential for diversion to retailer $r$, which we label the ``RRC effect.''  Furthermore, the margin on product $w$ sold by retailer $r$ is potentially higher due to EDM, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.  These impacts tend to raise the resulting input price.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Simulations}
In this section, we use a similar setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations and allow for the presence integrated rival firms.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Generating Process}
Broadly speaking, we consider four categories of mergers: downstream, upstream, vertical, and integrated.  We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.  Similarly, upstream mergers are those between two unintegrated wholesalers or between an unintegrated wholesaler and a vertically integrated wholesaler/retailer.  We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.  Finally, we assume that an integrated merger is between two firms that are already vertically integrated pre-merger.

For downstream, upstream, and vertical mergers, we simulate markets by randomly sampling shares from a Dirichlet distribution for 2 to 5 retailers or wholesalers, respectively.\footnote{We parameterize the Dirichlet distribution so it is equivalent to a uniform distribution.  }  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers in our simulated markets to 7. We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers). Vertically integrated firms are not siloed: integrated wholesalers supply inputs to non-integrated retailers and vertically integrated retailers purchase inputs from non-integrated wholesalers.

@@

For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.  This results in 1.96 million possible merger simulations.  After eliminating situations where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, we have 1.37 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%. } Each market treats as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs, which we allow to be either constant or linear.

@@

The median average wholesale pre-merger price is \$4.60, and the median average retail pre-merger price is \$14.  Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.  Pre-merger HHIs range from 2,814 at the $25^{th}$ percentile to 5,030 at the $75^{th}$, with a median of 3,703. HHIs for vertical mergers increase by 1,277 points at the median, resulting in a median post-merger HHI equal to 4,889.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.  }  HHIs for upstream mergers increase by 2,004 points at the median, resulting in a median post-merger HHI equal to 5,231. HHIs for downstream mergers increase by 2,054 points at the median, resulting in a median post-merger HHI equal to 4,889.   HHIs for integrated mergers increase by 2,746 points at the median, resulting in a median post-merger HHI equal to 7,220. Many of these markets fall into the span designated by the DOJ/FTC Horizontal Merger Guidelines as ``Highly Concentrated Markets,'' with post-merger HHIs over 2,500 points and HHI Changes greater than 200 points.\footnote{HHI thresholds are discussed in the 2010 Horizontal Merger Guidelines, Section 5.3.  }

Figure \ref{fig:surplussum} summarizes our results for welfare.  Each of the four panels show how mergers impact surplus for a particular set of agents (consumers, retailers, wholesalers, or the entire market combined, respectively).  Surplus is reported as a percentage change of total pre-merger expenditure in the downstream market.  Each panel contains four box and whisker plots, with each plot corresponding to a different type of merger. The whiskers show the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median. Negative outcome values imply harm, and positive values imply benefits.

@@

Moving on t retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas downstream mergers, vertical mergers, and integrated mergers always benefit retailers, upstream mergers harm retailers in about 67\% of all simulations. Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, which dominates that from upstream mergers, but not the retailer surplus distribution from vertical mergers.

Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 77\% of all upstream mergers, 23\% of vertical mergers, 20\% of downstream mergers, and about 15\% of integrated mergers. %\textbf{comparing these results to \citeN{ST2021}, the effects are less stark (i.e. less of a transfer.}
However, only wholesaler surplus from upstream mergers first-order stochastically dominates the wholesaler surplus of the other merger types.

@@

Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.  Figure \ref{fig:CVvertincumbBW} depicts paired box and whisker plots summarizing consumer harm (blue/darker bars on the left in each pair) and total harm (orange/lighter bars on the right in each pair) as the number of incumbent integrated firms increases from 0 to 6. The plots when the number of  incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated. By contrast, for downstream and upstream  mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, with no integrated third parties. For integrated mergers, the plots when the number of incumbent integrated firms equals 2 (the lowest number allowed in this instance) depict a merger between two integrated firms, again with no integrated rivals.

We begin with vertical mergers, as they are our simplest case, lacking preexisting integration at either merging firm.  Absent incumbent integration (0 integrated firms), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial in total surplus.  The addition of a single rival integrated firm (denoted as 1 integrated firm in the figure) narrows the distribution of outcomes slightly, but leaves median harm steady. Adding more integrated rivals tends to shift some outcomes from the lower tail to the upper tail, increasing the proportion of mergers with benefits, especially for consumers.  However, the interquartile range and median are largely unchanged.  Thus, the presence of rival integrated firms can limit the extent of harm, particularly consumer harm, from vertical mergers, though not dramatically.  Integrated rivals have sources of bargaining leverage that mirror the additional leverage a vertical merger brings to the combining firms.  It appears that these rivals can therefore somewhat limit the extent of negative impacts through RRC.

By contrast, not only are upstream mergers absent incumbent vertical integration rarely beneficial, but median consumer harm  from upstream mergers is about 6\% of pre-merger revenues  and total harm is about 3\% of pre-merger revenues. A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the proportion of beneficial mergers but does little to change median harm: consumer welfare increases in almost 16\% of mergers and total welfare increases in 4.3\% of mergers, but median consumer harm increases to 6.3\% and median total harm increases to 4.8\% of pre-merger revenues. Increasing the number of integrated 3rd parties to 1 (2 integrated firms) disproportionately shrinks the inter-quartile range-- relatively more harmful mergers are eliminated than beneficial ones-- but the presence of a single integrated 3rd party has little impact on median harm. Adding additional integrated 3rd parties (3,4 integrated firms) further decreases the range of outcomes while leaving median harm largely unchanged. In summary, additional incumbent integrated firms do not substantially impact median harm, but do substantially reduce the likelihood of observing extremely harmful mergers.

Like upstream mergers, downstream mergers absent incumbent vertical integration are rarely beneficial, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues while total harm is about 5\% of pre-merger revenues. A merger between an incumbent integrated firm and an unintegrated retailer when all other market participants are unintegrated (1 integrated firm)  increases the range of outcomes while also increasing median harm: consumer welfare increases in about 18\% of mergers and total welfare increases in 5.7\%, but median consumer harm increases to 14\% and median total harm increases to 7.3\% of pre-merger revenues. Increasing the number of integrated 3rd parties to 1 (2 integrated firms) shrinks the inter-quartile range but has little impact on median harm. Adding additional integrated 3rd parties (3 integrated firms) further decreases the range of outcomes while also decreasing median harm t 5.6\%. With 3 integrated 3rd parties (4 integrated firms), a downstream merger between an integrated and unintegrated firms benefits consumers in about 27\% of markets and shrinks median consumer harm to 6.4\% of pre-merger revenues. By contrast, with 3 integrated 3rd parties, the inter-quartile range of total harm shrinks, and median total harm falls to about 3.8\% of pre-merger revenues. These results indicate that unlike either vertical upstream mergers, integrated downstream mergers are more harmful than unintegrated downstream mergers, and that the presence of additional integrated rival incumbents reduces harm, though slowly.

Finally, absent the presence of rival incumbent integrated firms, mergers between two integrated firms benefits consumers in about 7\% of simulated markets and is net beneficial in 15\% of simulated markets. While adding one additional rival incumbent integrated firm (3 integrated firms) has little effect on either the distribution of either outcome, adding a second integrated rival (4 integrated firms) eliminates the most harmful integrated mergers while also increasing median consumer harm to 14\%  and median total harm to 10.6\%. The presence of additional integrated incumbents has little incremental effect on  median harm.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mergers and Bargaining Power}
\citeN{ST2021} show using numerical simulation that downstream and vertical mergers where wholesalers have relatively more bargaining power are less harmful than mergers where retailers have relatively more bargaining power. Here, we show that this result continues to hold when there are rival incumbent integrated firms. We also show that this relationship holds for integrated mergers as well as downstream and upstream mergers when one of the merging parties is vertically integrated.

%Figures \ref{fig:CVbargupBW}-\ref{fig:CVbargbothBW}  depicts box and whisker plots summarizing the consumer, retailer, wholesaler and total welfare effects for each merger as the number of incumbent integrated firms increases. For upstream, downstream and vertical mergers, the figures depict box and whisker plots for 0,1 and 4 incumbent vertically integrated firms.  For downstream and upstream  mergers (Figures \ref{fig:CVbargupBW} and \ref{fig:CVbargdownBW}), the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated. By contrast, for vertical mergers ((Figures \ref{fig:CVbargvertBW}), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated. Finally, Figure \ref{fig:CVbargbothBW} depicts box and whisker plots for integrated mergers with 2,4, or 6 integrated firms.  When the number of  incumbent integrated firms equals 2, Figure \ref{fig:CVbargbothBW} depicts a merger between two integrated firms, with all rivals are unintegrated. Plots with either 4 or 6 integrated firms depict mergers between integrated incumbents when there are either 2 or 4 rival integrated firms in the market.

Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top) and total (bottom) welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter increases from 1/9 (retailers have the advantage to 9 (wholesalers have the advantage). also depicted are three sets of box and whisker plots that correspond to the number of incumbent integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) incumbent vertically integrated firms.  As in Figure \ref{fig:CVvertincumbBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated: these are comparable to those in \citeN{ST2021}. For vertical mergers (left), the plots when the number of  incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when a single 3rd party is integrated. By contrast, for upstream (middle) and downstream (left) mergers, the plots when the number of  incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated rival, and all 3rd parties are unintegrated. Finally, plots with 4 integrated firms increases the number of rival incumbent integrated firms in upstream and downstream mergers to 3 and in vertical mergers to 4 .

For vertical mergers, the presence of integrated incumbents has little impact on the strong, positive relationship between bargaining power and harm for vertical mergers. Figure \ref{fig:CVbargnobothBW} shows that consumer harm when there are fewer incumbent integrated firms first-order stochastically dominates consumer harm when there are more incumbent integrated firms, the difference is smallest when wholesalers and retailers have equal bargaining power and greatest as wholesaler power increases relative to retailer power. Across all three incumbent integrated firm scenarios, median consumer harm decreases from about 7\% when retailers have the advantage to -24\% when wholesalers have the advantage. Similarly, median total harm decreases from about 2.9\% of pre-merger expenditures when retailers have relatively more bargaining power to about -14\% of pre-merger expenditures when wholesalers have relatively more bargaining power.

Turning to upstream mergers, when no incumbent firms are in the market (0 integrated firms), there is a positive relationship between bargaining power and both consumer and total harm.  Median consumer harm increases from 2.2\% of pre-merger expenditures when retailers have relatively more bargaining power to about 29\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from 0.7\% of pre-merger expenditures when retailers have relatively more bargaining power to about 12\% when wholesalers have relatively more bargaining power. By contrast, for mergers between an integrated and unintegrated upstream supplier (1 integrated firm), there is a positive relationship between bargaining power and both consumer and total surplus. Median consumer harm decreases from about 8.4\% of pre-merger expenditures when retailers have relatively more bargaining power to about -24\% of pre-merger expenditures when wholesalers have relatively more bargaining power. Likewise, median total harm increases from about 3.5\% of pre-merger expenditures when retailers have relatively more bargaining power to -8.4\% when wholesalers have relatively more bargaining power. Adding 2 additional incumbent integrated firms (4 integrated firms) mitigates some of the adverse effects of mergers while still preserving the positive relationship between bargaining power and harm.

>>>>

\begin{center}
Preliminary and Incomplete
\end{center}

@@

In this paper, we use a merger simulation model to assess the welfare implications of these complex mergers.  Our model is drawn directly from \citeN{ST2021}, and features a Bertrand logit downstream setup alongside a Nash Bargaining wholesale negotiation upstream.  We use this model to numerically simulate four types of mergers: (1) vertical mergers between an unintegrated retailer and an unintegrated wholesaler, (2) mixed downstream horizontal/vertical mergers between an unintegrated retailer and an integrated retailer/wholesaler pair, (3) mixed upstream horizontal/vertical mergers between an unintegrated wholesaler and an integrated retailer/wholesaler pair, and (4) integrated mergers between two previously integrated retailer/wholesaler pairs.  In each of these simulations, we calculate the impacts on consumer welfare and firm profits.  Furthermore, we examine how the presence of additional integrated rivals in the market changes our results. Our baseline simulations assume that units are produced with constant marginal costs, but we later extend our analysis to linear upward-sloping marginal costs.

%We find.

@@

%Our paper is related to the literature on mergers and vertical integration.

%The paper proceeds as follows.

@@

Turning to input prices, when wholesaler $w$ bargains with unaffiliated retailer $s$, the first order condition becomes
\begin{equation}
\begin{split}
&[p^W_{sw} - c^W_{sw}]s_{sw} - \sum_{t \in \mathbb{R}^w \setminus \{r,s\}} [p^W_{tw} - c^W_{tw}] \Delta s_{tw}(\mathbb{W}^s \setminus \{w\})\\
& - \overbrace{\overbrace{[p_{rw}-c^W_{rw}-c^R_{rw}] \Delta s_{rw}(\mathbb{W}^s \setminus \{w\})}^{\text{indirect EDM effect}} - \sum_{x \in \mathbb{W}^r \setminus \{w\}} [p_{rx} - p^W_{rx} - c^R_{rx}] \Delta s_{rx}(\mathbb{W}^s \setminus \{w\})}^{\text{RRC effect}}=\\
&\frac{1-\lambda}{\lambda} \left([p_{sw} - p^W_{sw} - c^R_{sw}]s_{sw} - \sum_{x \in \mathbb{W}^s \setminus \{w\}} [p_{sx} - p^W_{sx} - c^R_{sx}] \Delta s_{sx}(\mathbb{W}^s \setminus \{w\})\right).
\end{split}
\label{eq: vmerger wh upstream foc}
\end{equation}
which reflects the change in the disagreement payoff coming from the merger with retailer $r$.  Now when the wholesaler considers the possible loss of sales upon ceasing to trade with retailer $s$, these losses are softened due to a potential for diversion to retailer $r$, which we label the ``RRC effect.''  Furthermore, the margin on product $w$ sold by retailer $r$ is potentially higher due to EDM, as shown through the expression labeled ``indirect EDM effect,'' which can further compensate the firm.  These impacts tend to raise the resulting input price.

@@

%Integrated rivals.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Simulations}
In this section, we use a similar setup to that in \citeN{ST2021}, but allow for additional types of mergers beyond purely horizontal and purely vertical combinations and allow for the presence of integrated rival firms.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Generating Process}
Broadly speaking, we consider four categories of mergers: downstream, upstream, vertical, and integrated.  We define a downstream merger as a merger between two unintegrated retailers or between an unintegrated retailer and a vertically integrated wholesaler/retailer combination.  Similarly, upstream mergers are those between two unintegrated wholesalers or between an unintegrated wholesaler and a vertically integrated wholesaler/retailer.  We define vertical mergers as those between an unintegrated wholesaler and unintegrated retailer.  Finally, we assume that an integrated merger is between two firms that are both already vertically integrated pre-merger.

For downstream, upstream, and vertical mergers, we simulate markets by randomly sampling shares from a Dirichlet distribution for 2 to 5 retailers or wholesalers, respectively.\footnote{We parameterize the Dirichlet distribution so it is equivalent to a uniform distribution.  }  Because integrated mergers must have two vertically integrated incumbents in the pre-merger state, for those simulations we increase the maximum allowable number of wholesalers and retailers in our simulated markets to 7. We also assume that in the pre-merger state, there are anywhere from 0 to 4 vertically integrated incumbents (6 for integrated mergers). Vertically integrated firms are not siloed: integrated wholesalers supply inputs to retailers other than their integrated partner and vertically integrated retailers purchase inputs from wholesalers other than their integrated partner.

@@

For each combination of number of retailers, number of wholesalers, number of incumbent integrated firms, and bargaining parameter, we draw 1,000 different sets of market primitives.  After eliminating situations where the merger is unprofitable to the merging firms, as well as markets that do not pass the Hypothetical Monopolist Test, we have 2.7 million markets remaining.\footnote{The Hypothetical Monopolist Test is a market definition exercise that checks whether a monopolist that owns all products in a candidate market would raise the price of at least one of the merging producers' products by at least a ``small but significant non-transitory increase in price'' (SSNIP), which we take to be 5\%. } Each market treats as primitives the number of retailers, the number of wholesalers, the bargaining parameter, and the wholesaler and retailer marginal costs.

@@

The median average wholesale pre-merger price is \$4.50, and the median average retail pre-merger price is \$13.  Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.  Pre-merger HHIs range from 2,814 at the $25^{th}$ percentile to 5,030 at the $75^{th}$, with a median of 3,703. HHIs for vertical mergers increase by 1,298 points at the median, resulting in a median post-merger HHI equal to 5,995.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.  }  HHIs for downstream mergers increase by 2,015 points at the median, resulting in a median post-merger HHI equal to 4,914.  HHIs for upstream mergers increase by 2,063 points at the median, resulting in a median post-merger HHI equal to 5,238. HHIs for integrated mergers increase by 2,770 points at the median, resulting in a median post-merger HHI equal to 7,280.

Figure \ref{fig:surplussum} summarizes our results for welfare.  Each of the four panels show how mergers impact surplus for a particular set of agents (consumers, retailers, wholesalers, or the entire market combined, respectively).  Surplus is reported as a percentage change of total pre-merger expenditures in the downstream market.  Each panel contains four box and whisker plots, with each plot corresponding to a different type of merger. The whiskers show the $5^{th}$ and $95^{th}$ percentiles of the outcome distribution, the boxes denote the $25^{th}$ and $75^{th}$ percentiles, and the solid horizontal line marks the median. Negative outcome values imply harm, and positive values imply benefits.

@@

Moving on to retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas downstream mergers, vertical mergers, and integrated mergers always benefit retailers, upstream mergers harm retailers in about 67\% of all simulations. Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, which dominates that from upstream mergers, but not the retailer surplus distribution from vertical mergers.

Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 77\% of all upstream mergers, 23\% of vertical mergers, 20\% of downstream mergers, and about 15\% of integrated mergers. However, only wholesaler surplus from upstream mergers first-order stochastically dominates the wholesaler surplus of the other merger types.

@@

Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.  Figure \ref{fig:CVvertincumbBW} depicts paired box and whisker plots summarizing consumer harm (blue/darker bars on the left in each pair) and total harm (orange/lighter bars on the right in each pair) as the number of incumbent integrated firms increases. The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated. By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties. For integrated mergers, the plots when the number of incumbent integrated firms equals 2 (the lowest number allowed in this instance) depict a merger between two integrated firms, again with no integrated rivals.

We begin with vertical mergers, as they are our simplest case, because there is never any preexisting integration at either merging firm.  Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW}), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial in total surplus.  The addition of a single rival integrated firm (1 integrated firm in the figure) narrows the distribution of outcomes slightly, but leaves median harm basically the same. Adding more integrated rivals tends to shift some outcomes from the lower tail to the upper tail, increasing the proportion of mergers with benefits, especially for consumers.  However, the inter-quartile range and median are largely unchanged.  Thus, the presence of rival integrated firms can limit harm, particularly consumer harm, from vertical mergers, though not dramatically.  Integrated rivals have sources of bargaining leverage that mirror the additional leverage a vertical merger brings to the combining firms.  It appears that these rivals can therefore somewhat limit the extent of negative impacts that would otherwise occur through RRC.

Moving to the left-most panel of Figure \ref{fig:CVvertincumbBW}, downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues and total harm equal to about 5\% of pre-merger revenues. \citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.  Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 18\% of mergers, and total welfare increases in 5.7\%, but median consumer harm grows to 14\% of pre-merger revenues, and median total harm grows to 7.3\%. Once one of the merging firms is already integrated, the merger now has potential EDM and RRC effects, which raises the possibility of both benefits and harms to welfare.  On net, we find that the harms dominate in most instances.  Adding integrated third party rivals (2-4 integrated firms in the figure) somewhat decreases the range of outcomes and moves the distribution towards less harm.  With 3 integrated third parties (4 integrated firms in the figure), the merger benefits consumers in about 27\% of markets and shrinks median consumer harm to 6.4\% of pre-merger revenues. The inter-quartile range of total harm narrows, and median total harm falls to about 3.8\% of pre-merger revenues. As we saw with vertical mergers, it appears that the presence of third party integrated rivals can limit harm, though here the effect is a bit stronger.

Next we examine the welfare impacts of upstream mergers.  Like with downstream mergers, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues. This result is to be expected, as upstream mergers without additional integration do not have the potential countervailing effects for welfare that have been discussed in the context of downstream or vertical mergers.  Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear, but median harm is largely unchanged.  Consumer welfare increases in almost 16\% of mergers and total welfare increases in about 4.3\% of mergers, but median consumer harm and total harm are roughly steady at approximately 6.3\% and 4.8\% of pre-merger revenues, respectively.  The additional vertical integration creates some opportunities for EDM to enhance welfare, but this must be balanced against the potential for RRC.  Just as we saw with downstream mergers, here we find that harm dominates in most simulations.  Adding in integration by rival firms (2-4 integrated firms in the figure) narrows the inter-quartile range and trims the lower whisker-- meaning that relatively more harmful mergers are eliminated than beneficial ones-- and slightly shrinks median harm.  Therefore, the presence of third-party incumbent integrated firms, as in the vertical and downstream cases, can limit the likelihood of observing extremely harmful mergers.

We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW}.  Absent the presence of rival incumbent integrated firms (2 integrated firms in the figure), mergers between two integrated firms benefits consumers in about 7\% of simulated markets and are net beneficial in 15\% of simulated markets. Thus, we find that mergers between two firms that are already integrated are harmful in the vast majority of cases.  While adding one additional rival incumbent integrated firm (3 integrated firms in the figure) has little effect on the distribution of either consumer or total welfare, adding a second integrated rival (4 integrated firms in the figure) eliminates the most harmful integrated mergers by shortening the lower whisker. Adding further integrated incumbents has little incremental effect on the distribution of harm.

To summarize, our analysis of Figure \ref{fig:CVvertincumbBW} yields some overall conclusions.  First, we find that mergers where one or both of the merging firms are already integrated (and no rivals are integrated) are only beneficial in a small proportion of instances.  Second, the presence of integrated rivals can decrease the likelihood of observing the most harmful mergers, though the strength of that effect can vary widely across types of mergers.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mergers and Bargaining Power}
\citeN{ST2021} show using numerical simulations that downstream and vertical mergers when wholesalers have relatively more bargaining power are less harmful compared to when retailers have relatively more bargaining power. Here, we find that this relationship also holds for downstream and upstream mergers when one of the merging parties is already vertically integrated and for integrated mergers.  We also show that the result persists for all types of mergers when there are rival non-merging integrated firms.

Figure \ref{fig:CVbargnobothBW}  depicts box and whisker plots summarizing the consumer (top panels) and total (bottom panels) welfare effects for downstream, upstream, and vertical mergers as the relative bargaining power parameter goes from 9 (wholesalers have the advantage) to 1/9 (retailers have the advantage). Also depicted for each bargaining power parameter are three sets of box and whisker plots that correspond to the number of incumbent vertically integrated firms included in the simulated markets: 0 (light blue), 1 (darker blue), and 4 (darkest blue) firms.  As in Figure \ref{fig:CVvertincumbBW}, plots when the number of integrated firms is 0 assume that pre-merger, no firms in the market are vertically integrated; these are comparable to those in \citeN{ST2021}. For vertical mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome from an unintegrated wholesaler merging with an unintegrated retailer when one third party rival is integrated. By contrast, for upstream and downstream mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated firm when all third parties are unintegrated. In turn, plots with 4 integrated firms increase the number of rival incumbent integrated firms in vertical mergers to 4 and in upstream and downstream mergers to 3.  Figure \ref{fig:CVbargbothBW} provides analogous results for integrated mergers.  In integrated mergers, when the number of  incumbent integrated firms equals 2, this depicts a merger between two integrated firms with all rivals are unintegrated. Plots with either 4 or 6 integrated firms depict mergers between integrated incumbents when there are either 2 or 4 rival integrated firms in the market, respectively.

We begin by examining vertical mergers.  We find that the presence of integrated incumbents has little impact on the relationship between relative bargaining power and welfare. Consistent with the patterns seen in \citeN{ST2021}, mergers tend to benefit welfare when wholesalers have relatively more power, and harm welfare when retailers have relatively more power.  Larger wholesaler bargaining power offers more possibilities for EDM, as pre-merger input prices are likely to be high.  This channel becomes less relevant as retailers gain more power.  The impact on consumers is roughly zero when wholesalers and retailers have equal power.  Furthermore, we find that consumer harm when there are fewer incumbent integrated firms first-order stochastically dominates harm when there are more incumbent integrated firms, although the magnitude of these differences is small. Across all three integrated incumbent scenarios, median consumer harm moves from about -24\% of pre-merger expenditures when wholesalers have the advantage to 7\% when retailers have the advantage.  Similarly, median total harm goes from from about -14\% of pre-merger expenditures when wholesales have the advantage to 2.9\% when retailers have the advantage.

For downstream mergers, we continue to find that higher retailer relative bargaining power leads to more negative impacts on welfare.  Absent any incumbent integration (0 integrated firms), all the outcomes shown in Figure \ref{fig:CVbargnobothBW} feature a decrease in welfare, which mirrors our finding in Figure \ref{fig:CVvertincumbBW}.  Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 18\% of pre-merger expenditures when retailers have relatively more bargaining power. Likewise, median total harm moves from about 2\% of pre-merger expenditures to 7.6\%.  In situations where retailers already have more bargaining power, retailers are likely to have extracted significant surplus from wholesalers prior to the merger, which limits any potential benefits from increased bargaining leverage.  Thus, there is little to counteract harms from decreased downstream competition.
